---
title: "Week 4a Pre-lecture exercises"
author: ''
date: ""
output:
  pdf_document:
    fig_caption: yes
  html_document:
    df_print: paged
geometry: margin=1in
fontsize: 10pt
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align="center", fig.height=4.5, fig.width=5, collapse=TRUE, comment="", prompt=FALSE, echo = TRUE, cache=TRUE, autodep=TRUE, tidy=FALSE, tidy.opts=list(width.cutoff=50))
options(width=75)
```



Please watch the videos from the *Week4a* playlist under the *Panopto Video* tab on Canvas.  
Remember that you can adjust the speed and the size of the videos too. Please knit the .rmd file and submit your solutions in a .pdf file on Canvas.

You can see the *Examples of different R objects* section in Chapter 2.3 if you would like to read more about R object types.  The *Matrix Plot* section in the same chapter also gives an example for making a matrix plot.

1. We have an R object, and when we print the object in the console, the output is the following (the first column is just for the row indices, so the column is not part of the object): 

```
  Sepal.Length Sepal.Width Petal.Length Petal.Width Species
1          5.1         3.5          1.4         0.2  setosa
2          4.9         3.0          1.4         0.2  setosa
3          4.7         3.2          1.3         0.2  setosa
4          4.6         3.1          1.5         0.2  setosa
5          5.0         3.6          1.4         0.2  setosa
6          5.4         3.9          1.7         0.4  setosa
```

What kind of R object is this?  There should be more than one correct answer to this question.

> **This is a data frame and an array of decimals/doubles (cols 1-4) and strings (col 5)**

2. Create a list 'l' with 2 elements.  The 2 elements' names should be `num` and `mat`.  The first element is a vector of a sequence of numbers starting from 8, ending at -28 with decrement 3.  The second element is a 3 x 3 matrix; please put in whatever numbers you like in the matrix.

```{r question-2}
  l = list(num = seq(from = 8, to = -28, by = -3),
           mat = matrix(1:9, nrow = 3, ncol = 3))
l
```



3. (This is also the last question on Precept 4)

**Part a**

Use the dataset "diamonds" from the {ggplot2} package. Compare the pairwise relationship between the variables: `price`, `carat`, `depth` and `table` for different categories in `cut`.  Use `ggpairs()` to do this.  For all the continuous variable pairs:  Display the correlation coefficients on the lower-triangular plots, and display the scatterplots on the upper-triangular plots. 
 
Reminder 1: For all graphs that you produce, there should be no ambiguity and all the labels should be legible.  
 
Reminder 2: You can adjust the size of the graphs produced in a particular code chunk with the `fig.height` and `fig.width` arguments as in ```{r fig.height=8}```
 
```{r question-3}
#install.packages("ggplot2", dependencies = T)
#install.packages("GGally", dependencies = T)
#install.packages("dplyr", dependencies = T)
#library(ggplot2)
#library(GGally)
#library(dplyr)

pairs_diamonds <- diamonds |>
  group_by(cut) |>
  select(price, carat, depth, table)
ggpairs(pairs_diamonds,
        mapping = aes(color = factor(cut), alpha = 0.4),
        columns = 1:ncol(pairs_diamonds),
        title = "Pairwise Relationships for Price, Carat, Depth, and Table Within\n Diamonds Grouped by Cut",
        lower = list(continuous = "cor"),
        upper = list(continuous = "points"))
```
 


**Part b**

Based on your graph in part a, is there a strong linear relationship between the variables `price` and `depth`?  What about between `price` and `carat`?

> **No, there is a relatively weak linear relationship between price and depth because the points on the scatterplot aren't as tightly clustered together as they could be and they're on a straight line going vertically upwards. Furthermore, the correlation is low and nowhere close to +1, and cut quality is poor in comparison to other diamonds. There is a strong linear relationship between price and carat because the points demonstrate an upwards linear trend even though they aren't densely clustered together. Even still, correlation is high and close to 1 which suggests that there is a strong linear relationship.**

What is the correlation coefficient for the `price` and `carat` among diamonds with very good cut quality?

> **The correlation coefficient for price and carat is 0.922 which is close to +1 and they have a high cut quality across all categories. More specifically, among diamonds with very good cut quality, price and carat are the highest.**
