---
title: "Week 8 Pre-lecture exercises"
author: 
subtitle: ''
date: ""
output:
  pdf_document:
    fig_caption: yes
  html_document:
    df_print: paged
geometry: margin=1in
fontsize: 10pt
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align="center", fig.height=4.5, fig.width=5, collapse=TRUE, comment="", prompt=FALSE, echo = TRUE, cache=TRUE, autodep=TRUE, tidy=F, tidy.opts=list(width.cutoff=50))
options(width=75)
```

Remember that you can adjust the speed and the size of the videos.  Please knit the .rmd file and **submit your solutions in a .pdf** file on Canvas.

Please watch the videos from the *Week8* playlist under the *Panopto Video* tab on Canvas. 



1.  Let $Z$ be a standard Normal random variable, so $Z$ follows the `Normal(0, 1)` distribution.  To generate a number for $Z$ means to generate  a number according to the probability density curve for the `Normal(0, 1)` distribution.  For the `Normal(0, 1)` density curve, we note that the area under the density curve becomes smaller for values further away from zero.  This means that if we generate numbers for $Z$, the numbers that are closer to zero are more likely to show up than the numbers that are further away from zero.  We can think of generating 100 values for $Z$ as drawing a random sample of size 100 with replacement from the `Normal(0, 1)` distribution.


(a.) Generate 500 random numbers from the distribution `Normal(0, 1)` and assign the vector that stores these numbers to the variable `small.sample`.  Make a histogram for the data in `small.sample`.  Superimpose the density curve for `small.sample` and that for `Normal(0, 1)` on the histogram.  Set seed for the simulation to be `66943`.  Recall that we can adjust the thickness of the lines with the input argument `linewidth` for any geom functions that draw lines.  How close do the two density curves look?  


```{r}
library(ggplot2)
set.seed(66943)
small.sample = rnorm(500)

df.d <- tibble::tibble(dat = small.sample)
ggplot(df.d, mapping = aes(x = dat)) + 
  geom_histogram(mapping = aes(y = after_stat(density)), color = "white", fill = "steelblue") +
  geom_density(aes(color = "Sample data"),
               linewidth = 1.2) +
  stat_function(fun = dnorm, 
                args = list(mean = mean(df.d$dat), sd = sd(df.d$dat)), 
                mapping = aes(color = "Normal distribution"), 
                linewidth = 1.2) +
  scale_colour_manual("Density Curve", 
                      values = c("Normal distribution" ="pink","Sample data" = "yellow"))
```

> The density curves look very close and the histogram for the data in small.sample is symmetric and in a bell shape which shows that the small.sample data is from a normal distribution.

(b.) For the numbers in `small.sample`, what fraction of them are less and equal to 1.3?  Recall that $Z$ is a standard Normal random variable.  What is $P(Z \leq 1.3)$?  Compare the two numbers.

```{r}
mean(small.sample <= 1.3)
pnorm(1.3)
```

> The two values are approximately the same and of the same magnitude.

(c.) What is the 0.7 quantile of the data in `small.sample`?  Compare this number to the 0.7 quantile of the standard Normal distribution.

```{r}
quantile(small.sample, 0.7)
qnorm(0.7)
```

> The two values are approximately the same and of the same magnitude.

(d.) For the numbers in `small.sample`, what fraction of them are between -1.1 and 0.7, inclusively?   Find $P(-1.1 \leq Z \leq 0.7)$.

```{r}
mean(small.sample >= -1.1 & small.sample <= 0.7)
pnorm(0.7) - pnorm(-1.1)
```

(e.) Find $P(Z \leq -1.1$ or $Z \geq 0.7)$.

```{r}
# Formula for the union rule: 
# P(Z <= -1.1 U Z >= 0.7) = P(Z <= -1.1) + P(Z >= 0.7) - P(Z <= -1.1 & Z >= 0.7)
pnorm(-1.1) + (1 - pnorm(0.7)) - 0
```




