---
title: "Precept 7 Exercise"
author: "Your Name"
date: ""
output:
  pdf_document:
    fig_caption: yes
    number_sections: no
    toc: no
    toc_depth: 3
  html_document:
    df_print: paged
    toc: no
    toc_depth: 3
geometry: margin=1.5in
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align="center", fig.height=4, fig.width=6, collapse=TRUE, comment=NA, prompt=FALSE, echo = TRUE, cache=F, autodep=TRUE, tidy.opts=list(width.cutoff=67),tidy=F)
options(width=67)
```


# Objectives

* Learning functions: `dnorm(x, mean = 0, sd = 1, ...)`, `pnorm(q, mean = 0, sd = 1, ...)`, `qnorm(p, mean = 0, sd = 1, ...)`, `rnorm(n, mean = 0, sd = 1,)`;
* Constructing confidence intervals for different scenarios;
* Learn how to check Normality for a sample;
* CLT v.s. GCT



# Key Concepts

## Norm distribution

The following functions are related to X ~ Normal(mean, sd) random variables (the first letter of the function gives us a hint what the output is):

* dnorm(x, mean = 0, sd = 1, ...) -- returns **density** at X = x
* pnorm(q, mean = 0, sd = 1, ...) -- returns the **probablity** P(X <= q)  
* qnorm(p, mean = 0, sd = 1, ...) -- Given p, returns the **quantile** q such that the P(X <= q) = p 
* rnorm(n, mean = 0, sd = 1,) -- Generate a **random sample** of size `n`, drawn from the Normal(mean, sd) distribution

## Graphs for checking the Normality of a dataset

* geom_histogram()--checks the normality with a histogram
* stat_qq()--makes quantile-quantile plot against the N(0, 1) distribution
* stat_qq_line()--adds the line connecting the points at specified quartiles of the theoretical and sample distributions.
* stat_function(fun = dnorm, args = list(mean = ..., sd = ...)) --adds a Normal density curve for the Normal distribution with certain mean and sd.



## The Central Limit Theorem (CLT)


### CLT v.s. GCT


The vector `pop` contains the numbers in a population.

```{r}
# Data in population
pop = c(1,1,1,2,2,3)

mean(pop)
sd(pop)
```



```{r fig.height=8, echo = F}
# Make a function that generates 10000 samples;
# please do not worry about the code for this part
generator = function(pop, sample.size){
    return(sapply(1:10000, 
                  FUN=function(x){mean(
                    sample(pop, size=sample.size, replace=T)
                                      )}
                  )
           )
                                          }
```

The histogram for the population data is shown in the first graph below.

In the next three graphs, we observe the **sample mean** distribution when the sample size is 10, 100, and 1000 respectively.  **CLT** says that as the sample size gets larger and larger, the distribution of the **sample mean** gets closer and closer to the **Normal(mean=1.67, SD = $\frac{0.8165}{\sqrt{n}}$)** distribution, where $n$ is the sample size.

```{r fig.height=8, echo = F}
# Demo for CLT
par(mfrow=c(4,1))
hist(pop, freq = F, breaks=c(0.5,1.5,2.5,3.5), xlim = c(0.5, 3.5), 
     main = 'Population distribution')

set.seed(2048)
X=generator(pop,10)
hist(X,main='Sample mean distribution; sample size=10',
     xlim = c(0.5, 3.5),
     xlab = 'sample means', freq = F)
X=generator(pop,100)
hist(X,main='Sample mean distribution; sample size=100',
     xlim = c(0.5, 3.5),
     xlab = 'sample means', freq = F)
X=generator(pop,1000)
hist(X,main='Sample mean distribution; sample size=1000',
     xlim = c(0.5, 3.5),
     xlab = 'sample means', freq = F)
```

The **Glivenko–Cantelli theorem** (GCT) says that suppose you draw a sample with replacement from a population, as your sample size gets larger and larger the **histogram for your sample** will look more and more **similar** to the **histogram for the population** that you are drawing from.  

Note that the GCT is about the distribution of data in the sample, whereas the CLT is about the sample mean distribution.

```{r fig.height=8, echo = F}
# Demo for GCT
par(mfrow=c(4,1), mar=c(2, 2, 2, 2))
hist(pop, freq = F, breaks=c(0.5,1.5,2.5,3.5), main = 'Population distribution', ylim = c(0, 0.5))

Y=sample(pop,size=10,replace=T)
hist(Y,breaks=c(0.5,1.5,2.5,3.5), freq=F,
     main = 'sample size = 10 for one sample', xlab = 'data points in the sample', ylim = c(0, 0.5))
Y=sample(pop,size=100,replace=T)
hist(Y,breaks=c(0.5,1.5,2.5,3.5), freq=F,
     main = 'sample size = 100 for one sample', xlab = 'data points in the sample', ylim = c(0, 0.5))
Y=sample(pop,size=1000,replace=T)
hist(Y,breaks=c(0.5,1.5,2.5,3.5), freq=F,
     main = 'sample size = 1000 for one sample', xlab = 'data points in the sample', ylim = c(0, 0.5))

```


## Constructing confidence Intervals

### $(1-\alpha)$ confidence Intervals for estimating a population mean

* **Case 1: When the sample size $n$ is large:** The standardized sample means form a Normal(0, 1) approximately.  If the population SD $\sigma$ is known we can calculate the $SE(\bar X) = \frac{\sigma}{\sqrt{n}}$ ; if the population SD $\sigma$ is unknown, because the sample size is large, the sample SD will be a good estimate of the population SD, $\sigma$. **We use the Normal(0, 1) distribution to find out the value for $q_{(1-\alpha + \frac{\alpha}{2})}$.**

$$
\bar x \pm \ q_{(1-\alpha + \frac{\alpha}{2})} \frac{\sigma}{\sqrt{n} }  
$$
or

$$
\bar x \pm \ q_{(1-\alpha + \frac{\alpha}{2})} \frac{s}{\sqrt{n} }
$$

--------

* **Case 2: When the sample size is small, $\sigma$ is unknown but the *population is normal*:** 

If the population SD $\sigma$ is unknown to us, with the condition that the *population is normal* (or at least close to being symmetric) the standardized sample means form a t-distribution with degree of freedom = (sample size -1) = $n$-1.  **In this case we will use the t-distribution with degree of freedom $n$-1 to find out the value for $q_{(1-\alpha + \frac{\alpha}{2})}$.**

$$
\bar x \pm \ q_{(1-\alpha + \frac{\alpha}{2})} \frac{s}{\sqrt{n} }  
$$

* **Case 3: When the sample size is small *and* the *population is very skewed*:**

In this case, we do not have enough information to construction a C.I. for the population mean.

--------

### Confidence Intervals for estimating other population quantities


For other population quantities if we do not know the distribution of the sample counterpart, *Bootstrap sampling* can be used to approximate the distribution of the sample counterpart.  We can then use this approximated distribution to construct a confidence interval to estimate the population quantity.


# Exercises

**Please fill in any missing axis labels and graph titles yourself.**


### Question 1

For the population mentioned in the Central Limit Theorem section of the key concept section above, the population mean and SD are

```{r}
mean(pop)
sd(pop) # Side note:  technically, because we are calculating the population mean, 
# not a sample mean, the SD should actually be 
# sd(pop)*sqrt(length(pop) - 1)/sqrt(length(pop)) = 0.745356
```

Let's draw a sample of size 60 from the population. 

```{r}
set.seed(4321)
sampl.60 = sample(pop, size = 60, replace = T)
```

Let's assume that **you do not know the population mean or the population SD**, but you have access to the sample data, particularly the sample mean and sample SD.


```{r}
mean(sampl.60)
sd(sampl.60)
```


---------

#### Part a

Below is the histogram (please don't worry about the code in the RMD file) for the sample data `sampl.60`.  Fill in the blanks for the following questions.

```{r echo = F}
hist(sampl.60, main = 'Histogram for data in a sample of size 60',
     freq = F, xlab = 'data in a sample of size 60',
     breaks=c(0.5,1.5,2.5,3.5))
```

The distribution for the data points in the sample is skewed, just like population distribution; this is a result of the Glivenko–Cantelli theorem (GCT).  

Even though the sample distribution is skewed, the distribution of the sample mean will still be roughly Normally distributed; this is a result of the Central Limit Theorem (CLT).  

The SD of the Sample Mean distribution should be about 0.73.  


#### Part b

Construct a 80% CI to predict the population mean with your sample data.

```{r}
#t.test(sampl.60, conf.level = 0.8)
c(mean(sampl.60) - qnorm(p = .9)*sd(sampl.60)/sqrt(60),
  mean(sampl.60) + qnorm(p = .9)*sd(sampl.60)/sqrt(60))

```


#### Part c (no action required)

Instead of 60 now draw a sample of size 6000 from the population and again let's assume again that you do not know the population mean or the population SD.  

```{r}
set.seed(4321)
sampl.6000 = sample(pop, size = 6000, replace = T)
mean(sampl.6000)
sd(sampl.6000)
```

#### Part d

Does the magnitude of the sample SD (sample SD refers to the SD of the data in the sample) change when the sample size is increased from 60 to 6000?  Does the magnitude of the estimated SD(sample mean) change when the sample size is increased from 60 to 6000?  Each set of these numbers reflects the result of which theorem? Explain.

> No, the magnitude of the sample SD doesn't change when the sample size is increased. This reflects the Glivenko–Cantelli Theorem (GCT). Yes, the magnitude of the estimated SD(sample mean) does change when the sample size is increased. This reflects the Central Limit Theorem (CLT).

#### Part e

Construct a 80% CI to predict the population mean.

```{r}
#t.test(sampl.6000, conf.level = 0.8)
c(mean(sampl.6000) - qnorm(p = .9)*sd(sampl.6000)/sqrt(6000),
  mean(sampl.6000) + qnorm(p = .9)*sd(sampl.6000)/sqrt(6000))

```

#### Part f

The length of the CI for the sample of size 60 is about 10 times of that for the sample of size 6000.  You don't have to write out the explanation for this part, but think about why this is expected.



### Question 2
(6.10 in OpenIntro) **Life rating in Greece**. Greece has faced a severe economic crisis since the end of 2009. A Gallup poll surveyed 1,000 randomly sampled Greeks in 2011 and found that 25% of them said they would rate their lives poorly enough to be considered suffering.


#### Part a
Describe the population parameter of interest. What is the value of the point estimate of this parameter?

> The population parameter is the percentage of Greeks who would rate their lives poorly enough to be considered suffering.


#### Part b

Which case (see the Key Concepts section) are we in?  Check if the conditions required for constructing a confidence interval based on these data are met.

> We're in Case 1 because the sample size n is large (n = 1000 >= 30)

#### Part c


Construct a 95% confidence interval for the proportion of Greeks who are suffering.  Note that if you have a sample of 0-1 numbers, the SD for the sample data can be obtained with this formula:

$Sample\_SD = \sqrt{p(1-p)}\sqrt{n/(n-1)}$ 

where $p$ is the proportion of 1's in the sample and $n$ is the sample size.  For example, for a sample of size 100 that has 40 1's and 60 0's the sample SD is

```{r}
sqrt(.40*(1-.4))*sqrt(100/(100-1))
```

We can verify this result by recreating the sample with 40 1's and 60 0's and calculate its SD:

```{r}
test.sampl = rep(0:1, times = c(60, 40))

sd(test.sampl)
```


Construct the 95% CI:

```{r}
c(.25 - qnorm(p = 0.975)*(sqrt(.25*(1-.25))*sqrt(1000/(1000-1)))/sqrt(1000),
  .25 + qnorm(p = 0.975)*(sqrt(.25*(1-.25))*sqrt(1000/(1000-1)))/sqrt(1000))
```


#### Part d


Without doing any calculations and with the same set of sample data, describe if the confidence interval would be shorter or longer if we decided to use a higher confidence level.

> The confidence interval would be longer if we decided to use a higher confidence level

#### Part e

Without doing any calculations, describe if the confidence interval would be shorter or longer if we used a larger sample.

> The confidence interval would be shorter if we used a larger sample.

### Question 3

This dataset \emph{Newcomb.RData} includes data on the amount of time (in seconds) taken for a beam of light to travel from the Naval Observatory in Washington D.C. to the Washington Monument and back, a distance of 7.44373km.   The data was collected in 1880 by Simon Newcomb (working with Ablert Michelson).

```{r}

Newcomb =data.frame(
    Time = c(2.4828e-05, 2.4826e-05, 2.4833e-05, 2.4824e-05, 
             2.4834e-05, 2.4756e-05, 2.4827e-05, 2.4816e-05, 2.484e-05, 2.4798e-05,
             2.4829e-05, 2.4822e-05, 2.4824e-05, 2.4821e-05, 2.4825e-05, 2.483e-05,
             2.4823e-05, 2.4829e-05, 2.4831e-05, 2.4819e-05, 2.4824e-05, 2.482e-05,
             2.4836e-05, 2.4832e-05, 2.4836e-05, 2.4828e-05, 2.4825e-05, 2.4821e-05,
             2.4828e-05, 2.4829e-05, 2.4837e-05, 2.4825e-05, 2.4828e-05, 2.4826e-05,
             2.483e-05, 2.4832e-05, 2.4836e-05, 2.4826e-05, 2.483e-05, 2.4822e-05,
             2.4836e-05, 2.4823e-05, 2.4827e-05, 2.4827e-05, 2.4828e-05, 2.4827e-05,
             2.4831e-05, 2.4827e-05, 2.4826e-05, 2.4833e-05, 2.4826e-05, 2.4832e-05,
             2.4832e-05, 2.4824e-05, 2.4839e-05, 2.4828e-05, 2.4824e-05, 2.4825e-05,
             2.4832e-05, 2.4825e-05, 2.4829e-05, 2.4827e-05, 2.4828e-05, 2.4829e-05,
             2.4816e-05, 2.4823e-05), 
    Series = as.factor(
        c(1, 1, 1, 1, 1, 
          1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2,
          2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3, 3, 3,
          3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3))) 
str(Newcomb)
head(Newcomb)
```



#### Part a

Use ggplot() to generate a histogram for the time data. Superimpose the density plots for the sample data and the Normal distribution with the same mean and SD on the histogram.  You can use the following labels:


```{r eval = F}
# DENSITY PLOT

ggplot(Newcomb, mapping = aes(x = Time)) +
  geom_histogram(mapping = aes(y = after_stat(density)), color = "white", fill = "steelblue") +
  geom_density(aes(color = "Sample data"),
               linewidth = 1.2) +
  stat_function(fun = dnorm, 
                args = list(mean = mean(Newcomb$Time), sd = sd(Newcomb$Time)), 
                mapping = aes(color = "Normal distribution"), 
                linewidth = 1.2) +
  scale_colour_manual("Density Curve", 
                      values = c("Normal distribution" ="pink","Sample data" = "yellow")) +
  labs(x = "Time (in seconds)", y = "Proportion per second",
       title = 'Distribution for the Amount of Time (in seconds) 
       Taken for a Beam of Light to Travel',
     subtitle = 'from the Naval Observatory to the Washington Monument and Back')
  

```






#### Part b 

We already saw on the histogram that the sample distribution has some outliers with small values.  Use a qq-plot to compare the quantile of the data with the quantile from N(0, 1) to further confirm that the distribution of time is not normally distributed because of the outliners.  

```{r}
ggplot(Newcomb) +
  stat_qq(mapping = aes(sample = Time)) +
  stat_qq_line(mapping = aes(sample = Time), color = "red") +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles",
       title = "Normal Q-Q Plot for Newcomb")
  
```



#### Part c

Determine the 90\% confidence interval for the average time.  To construct your CI in a correct way does the population data need to be Normal?  

```{r}
c(mean(Newcomb$Time) - qnorm(p = 0.95)*sd(Newcomb$Time)/sqrt(length(Newcomb)),
  mean(Newcomb$Time) + qnorm(p = 0.95)*sd(Newcomb$Time)/sqrt(length(Newcomb)))
```

> No, the population doesn't need to be Normal as justified by CLT. For large n, you can use the Normal-based confidence interval since the distribution of the sample mean is approximately Normal.

#### Part d (In case Tuesday's lecture does not have enough time to cover this part, you are not required to complete Parts d and e to get the credit for this precept.  You can either complete these two parts on your own after Thursday's lecture or just do them next week--they will become the required questions for next week's precept.)

We would like to construct a 90\% confidence intervals to predict the population standard deviation of *Time*.  Since we do not know what kind of distribution the sample SD follows and how it relates to the population SD, we will use Bootstrap sampling to help us get an idea of the distribution of the sample SD.  Use bootstrap sampling with 20,000 simulations to determine the 90\% confidence intervals for predicting the population standard deviation of *Time*.  Use `set.seed(672)`.




#### Part e  (See comment in Part d)

Make a density plot of the bootstrap sample SD's above. Does the distribution of the bootstrap sample SD's look Normal?




