---
title: "Precept 2 Answers"
author: "Brittney Barnwell"
date: ""
output:
  pdf_document:
    fig_caption: yes
    toc: yes
  bookdown::html_document2:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  html_document:
    df_print: paged
    toc: yes
geometry: margin=1.5in
editor_options:
  chunk_output_type: console
---

## Please submit your solutions in a .pdf file for all your precepts from now on.


# Objectives

- Getting familiar with functions: `distinct()`, `arrange()`, `%in%` and `rank()`
- Learn how to create a report with the `{rmarkdown}` package
- Learn construction, extraction and assignment techniques for `list`s (and `data.frame`s)

# Rmarkdown Introduction

Rmarkdown is a **mark-up language** that can generate reports which include R code. As opposed to text processing programs that require marking the end-result in the document itself, this type of method marks the desired formatting and the final document is achieved by rendering the document. For example, *italicized text* can be marked with a single underscores or asterisks, **boldfaced text** is marked with double underscores or asterisks, header is marked with one or more `#` at the beginning of a line, etc.

The first thing you see in this type of file is the yaml header. The rendering process will use this to figure out the title and author name. You can also specify formatting option here such as, the type of output document (here a PDF instead of the default of HTML), include a table of content (`toc: true`), size of figures, adding captions to figure, etc.

# Knitr

To generate an output file with an .Rmd file, you need to render the .Rmd file using the button "Knit" that is on the top of the upper-left panel in RStudio. You can control the rendering options for the entire file using the following command, each option controls a different property in the file created:  

```{r setup, include=FALSE}
# This code chunk should usually be placed at the beginning of your report after the header
knitr::opts_chunk$set(fig.align="center", fig.height=5.5, fig.width=6, collapse=TRUE, comment='##', prompt=FALSE, echo = TRUE, cache=FALSE, autodep=TRUE, tidy.opts=list(width.cutoff=63), tidy=TRUE)
options(width=63)
```

# What the options in `knitr::opts_chunk$set()` mean

````
<!-- This is how you would comment in a .Rmd file -->
````

The options set in knitr::opts_chunk$set() control chunks in the entire report.

However, you can also use these options to control the settings for a particular code chunk; to do so you would put the option(s) inside the curly braces. E.G., all the graphs produced by the code in the following code chunk will be aligned left and will be 10 inches tall:


````
```{r fig.align="left", fig.height = 10}`r ''`
plot(x = 0:5, y = 10:15)
```
````


Below are the commonly used options and their descriptions, along with their default values and the data types of the input values in the parentheses.

**fig.align**: ('default'; character) alignment of figures in the output document (possible values are left, right and center; default is not to make any alignment adjustments).

**fig.height**: (7; numeric) Height in inches for plots created in chunk.

**fig.width**: (7; numeric) Width in inches for plots created in chunk.

**message**: (TRUE; logical) whether to preserve messages emitted by `message` (similar to warning).

**collapse**: (FALSE; logical; applies to Markdown output only) whether to, if possible, collapse all the source and output blocks from one code chunk into a single block.

**prompt**: (FALSE; logical) whether to add the prompt characters in the R code (see prompt and continue in ?base::options; note that adding prompts can make it difficult for readers to copy R code from the output, so prompt=FALSE may be a better choice.

**comment**: ('##'; character) the prefix to be put before source code output; default is to comment out the output by ##, which is good for readers to copy R source code since output is masked in comments (set comment=NA to disable this feature, though this is not recommended).

**cache**:  (FALSE; logical)  Whether to save results so that code blocks aren't re-run unless code changes,
_or_ a relevant earlier code block changed (see autodep).

**autodep**: (FALSE; logical) whether to figure out the dependencies among chunks automatically by analyzing the global variables in the code (may not be reliable) so that dependence does not need to be set explicitly.

**options(width=63)**: (80; numeric)  Controls the maximum number of columns on a line used in printing vectors, matrices and arrays.  


```{r load-packages}
library(babynames)
library(dplyr)
```

# Some code chunks for the demo

You can write **chunks** of code within the .Rmd file. This code can be incorporated as is, executed to include its output, or both. Below are a couple of examples for the use of code chunks using the `cars` dataset; the dataset lists the speed of 50 cars and the distances these 50 cars take to stop. We can see the size of the data frame using `dim`. We can use `head` to see that `cars` holds two `vector`s with numerical values. 

```{r cars-dim}
# within the code chunk everything is treated as if they were in a 
# .R script file; when you knit the .Rmd file all the code in the
# code chunks in the report will be run sequentially
# and lines of coed can be executed with ctrl+enter for PC and Command+Return for Mac
dim(cars)
```

```{r cars-head}
head(cars)
```

# `data.frames`

`data.frames` are rectangular stores of data made up of rows and columns, where each column can hold a different data type. They are the fundamental unit of analysis in modern R code and a large reason for R's success as a machine learning language. They should be familiar to people who have used spreadsheets or CSVs before.

For an example `data.frame`, we look at the `txhousing` data from the `{ggplot2}` package that we load using the `data` function.

```{r load-housing}
# getting the dataset `txhousing` from the `ggplot2` 
# package without loading the `ggplot2` package
data(txhousing, package='ggplot2')

txhousing %>% head() # `txhousing` is a tibble so it will prevent you from 
# printing the entire data frame in your report;
# in general, you should use `txhousing %>% head()` to 
# avoid output overflow
```

We check the class and shape of the data with `class()` and `dim()`.

```{r housing-info}
class(txhousing)
dim(txhousing)
```

We can also check the data types of the columns.

```{r}
# `strict.width` restricts the width of the output to avoid overflow
str(txhousing, strict.width = 'cut')
```

While we can use base-R functionality or a package like `{data.table}`, we'll use `{dplyr}` to perform most operations.

To extract a column as a `vector` we use `pull()`. But we should be careful, the data have `r nrow(txhousing)` so that would be a long `vector` to print. To avoid this we use `head()`, which works on both `data.frame`s and `vector`s as well as other types of objects.

```{r housing-vector}
txhousing %>% pull(sales) %>% class()
txhousing %>% pull(sales) %>% head()
```

Instead of `pull()` we use `select()` to choose columns. With `select()` no matter if we choose 1 or all of the columns, the result will still be a `data.frame`

```{r housing-select}
txhousing %>% select(sales) %>% head()
txhousing %>% select(sales) %>% class()
txhousing %>% select(sales) %>% dim()
txhousing %>% select(city, sales, volume, median) %>% head()
txhousing %>% select(city, sales, volume, median) %>% class()
txhousing %>% select(city, sales, volume, median) %>% dim()
```

# Some functions for manipulating data.frames: `distinct`, `arrange`

If we want to find the unique values of the `city` column in `txhousing` we use `distinct()` to find them and return the answer as a one-column `data.frame` with a row for each unique value.

```{r city-unique}
txhousing %>% distinct(city) %>% head()
```

We can use multiple columns to find jointly-unique combinations.

```{r city-year-unique}
txhousing %>% distinct(city, year) %>% head()
```

We sort an entire `data.frame` with `arrange()` and specifying which column to use for sorting.

```{r sales-sorts}
txhousing %>% arrange(sales) %>% head()
```

For descending sorting use the `desc()` function.

```{r sales-sorts-descending}

txhousing %>% arrange(desc(sales)) %>% head()

```

# Functions that are intrinsically for vector manipulations


## Summary statistics functions from the {base} package

Computing the numerical summary of data can be helpful in understanding the data.  Most of the summary Statistics functions that we have seen so far (e.g., `mean`, `sd`, `var`, `min`, `max`, `quantile`, `median` and `range`) are meant to be used on vectors.  For example, the following will work.

```{r}
mean(txhousing$sales, na.rm=TRUE)
```

However, the following will not work.

```{r}
mean(txhousing, na.rm=TRUE)
```

After all, when you have a data frame with elements of different data types, it would be difficult to interpret what you wanted to do if you asked `R` to take the mean of all the elements.

### Combining summary statistics functions with {dplyr} functions for data.frame manipulations

Combining summary statistics functions with `dplyr::summarize` allows us to compute numeric summaries for some columns of a `data.frame`. In case there are missing data we set `na.rm=TRUE` for functions like `mean` and `min`, etc.

```{r housing-summaries}
txhousing %>% summarize(AvgSale=mean(sales, na.rm=TRUE))
txhousing %>% 
  summarize(AvgSale=mean(sales, na.rm=TRUE), 
            Smallest=min(sales, na.rm=TRUE), 
            Largest=max(sales, na.rm=TRUE))
txhousing %>% summarize(SaleRange=range(sales, na.rm=TRUE))
```

Combining this with `dplyr::group_by` allows us to compute this information for subsets of the data.

```{r housing-summaries-grouped}
# calculate sales mean for each city
txhousing %>% 
  group_by(city) %>% 
  summarize(AvgSale=mean(sales, na.rm=TRUE))
```


```{r}
txhousing %>% 
  group_by(city) %>% 
  summarize(SaleRange=range(sales, na.rm=TRUE))
```

Note that because the output for `range()` has two elements, we have two rows for each city in the output data.frame above.  You need to watch out for unexpected results like this.  A better practice is to use

```{r}
txhousing %>% 
  group_by(city) %>% 
  summarize(Smallest=min(sales, na.rm=TRUE), 
            Largest=max(sales, na.rm=TRUE))
```


# Other functions for vector manipulations: `rank` and `%in%`

## `rank()` to get the ranks of the elements in a vector

`rank()` is used to rank elements in a vector in an increasinig order.

```{r}
rank(c(8, 10, 3, 1))
```

This is because 1 is the smallest number, 3 is the second smallest number, and so on.

### combine `rank()` with `dplyr::mutate()`

When combining with `dplyr::mutate()`, this allows us to have an output with a rank-column added to the selected columns from the original data.frame.  For example, to find how well each house performs in sales, we rank the `sales` column in an increasinig order using `rank()` (the column `city` is included so we can see which city the house is in):

```{r}
txhousing %>% select(city, sales) %>% mutate(SalesRank=rank(sales))

```

We can also sort the data by `sales` as seen earlier to see which city was on top. If we just want the top 5 performing cities, we use `slice_max`.

```{r sales-slice-max}
txhousing %>% select(city, sales) %>% slice_max(order_by=sales, n=5)
```

## `%in%` for value matching

For two *vectors* `x` and `y`, `x %in% y` returns a logical vector that have the same length as `x`.  The logical vector indicates whehter each element in `x` is in `y`.  For example,

```{r}
c(8, 19, 10, 432, 894, 32, 7, 10) %in% c(23, 53, 10, 32, 4)
```

### combine `%in%` with `dplyr::filter()`

It is convenient to combine `%in%` with `dplyr::filter()` to select rows in a data.frame that satisfy certain conditions.

```{r}
txhousing %>% select(city, sales) %>% filter(city %in% c('Tyler', 'Victoria'))
```

```{r}
max.sales = max(txhousing$sales, na.rm = T)
max.sales

txhousing %>% 
  select(city, sales) %>% 
  filter(sales  %in% max.sales)

```


**The following links is helpful for learning R Markdown (especially the first one).**

http://www.stat.cmu.edu/~cshalizi/rmarkdown/#mark-up-markdown

https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

https://yihui.name/knitr/options/

# Exercises 

Please include your answers in code chunks and make sure that you can knit your file successfully.

## Question 1

(a) Try placing one at a time of the following arguments into the curly brackets for the first 
code chunk and observe what happens when you compile (knit) the file. 

`include = FALSE` <!--evaluation but do not show codes or results-->  
`echo = FALSE`  <!--evaluate; do not show codes but show results-->  
`eval = FALSE`  <!--do not evaluate-->  

i.e., for the first one do this

```{r cars-dim-2, include = FALSE}
dim(cars)
```

```{r cars-dim-3}
knitr::opts_chunk$set(echo = FALSE)
dim(cars)
```

```{r cars-dim-3.5, echo = FALSE}
dim(cars)
```

```{r cars-dim-4, eval = FALSE}
dim(cars)
```
Side note: note that you can also add one of these arguments to `knitr::opts_chunk$set()`, and the option will then controls all the code chunk globally.


(b) Type your name for the `author` field in the header.

***Add a code chunk to each of the following questions to evaluate the answers***

## Question 2

`cars` is a built-in data frame in R.  

```{r show-cars}
class(cars)
dim(cars)
```

Without running any code, say what kind of R object each of the following line will return. Once you finish, verify your answers by using the `class` function. (You do not have to show the code to apply the `class` function; the verification is just for your own benefit.)

```{r cars-info, eval = F}
cars %>% pull(1) # type the R object kind here... 'vector'
cars %>% pull(speed) # ... 'vector'
cars %>% select(1) # type the R object kind here... 'data.frame'
cars %>% select(speed) # ... 'data.frame'
```




**Use the dataset `babynames` in the `{babynames}` package for the rest of the questions.**

## Question 3

Pick one person in your group.  For this person's name, we want to study its trend over the years, in terms of usage frequency. Extract the usage frequencies for this name.  Please keep the corresponding years too.

```{r name-freq}
babynames |> filter(name == "Minnie") |> select(n)
```


Just for your curiosity, you can see the trend by replacing `x_variable_vector` and `y_variable_vector` in the plot function below with the relevant vectors and then run the code (remember to remove `eval = F` if you want the graph to show up in your pdf).  Has the trend been increasing or decreasing since 1950?

```{r}

plot(x = babynames$year, y = babynames$n,
     xlab = 'years', ylab = 'Frequencies')

```
The trend was decreasing since 1950


## Question 4


(a)  Make a `data.frame` called `high_freq` by extracting the rows for female names that were used at least 20000 times in a given year between year 2000 and year 2017 inclusively.  
Hint: You should break down this problem into parts. Your thought/work process should follow steps similar to the following:

- Step 1: I am asked to extract the rows that satisfy the following conditions:
    i)  for female names
    ii)  for names used at least 20000 times in a given year
    iii)  for years between year 2000 and year 2017 inclusively
- Step 2: How do I extract rows that satisfy each of the conditions above?
- Step 3: Work out the code for Step 2 and test the code by checking the results produced by the code
- Step 4: How do I go from the results in step 3 to the requested result?

```{r baby-frequency}
high_freq = babynames |> filter(sex == "F", n >= 2000, year >= 2000, year <= 2017) |> group_by(name)
```

(b)  What are the names that appear in `high.freq`?  How many are there (do not include repeats)?

```{r frequent-names}
high_freq$name
nrow(distinct(high_freq))
```

(c)  Make a data frame `popular_names` by extracting the rows from `babynames` for the female names that you found in part (b) for years between 2000 and 2017, inclusively. Note that we want the rows in `babynames` that correspond to female names only. Compare the size of this data frame with that of `high_freq`.

```{r popular-names}
popular_names = data.frame(high_freq$name)
nrow(high_freq) > nrow(popular_names) | nrow(high_freq) < nrow(popular_names)
```


## Question 5: Rmarkdown error

Please remove `eval = F` for the following code chunk and knit your .rmd file to a .pdf file .  You will likely receive an error.  What does the error message say?  Can you guess what the problem is?

```{r}
library(babynames)
print.data.frame(babynames, max = 200000)
```
I didn't receive an error message but a lot of babynames were printed and it took a while for the page to load,
is this a problem?

The following is from the *Midterm 1 Practice Problem Set* that will be released later.  You might not have enough time to finish Part b.  Just try your best, and compare your answer with the answer key that will be posted later.


## Question 6: The Most popular babygirl names (Part of the midterm 1 practice problem set)

Suppose that most of you in this class are turning 19 to 22 this year; this means that you were born between years 2000 and 2003.  We will investigate the baby names in the dataset for this time period.  Please demonstrate that you know how to use pipes.

## Part a

Extract the rows in `babynames` that correspond to the names of the babies born between years 2000 and 2003, inclusively and name the resulting subset `names2000.2003`.  How many rows and columns does `names2000.2003` have? 


```{r}
names2000.2003 = babynames |> filter(year >= 2000, year <= 2003)
nrow(names2000.2003)
ncol(names2000.2003)
```

## Part b

Use `names2000.2003` from Part a for this Part. For the period between years 2000 to 2003 find out the top 20 most popular female names each year (it is okay to just print these four lists of names out in the code chunk; no need to include these names in the write-up).  Among these four lists of top 20 female names, which names stay in the top 20 at least twice for the time period 2000 to 2003?  

**Be careful**: If you decided to use the `arrange()` function, note that `arrange()` largely ignores grouping; you need to explicitly mention grouping variables and use .by_group = TRUE in order to group by them.  See the example using the `mtcars` dataset on the help page for `arrange()`.

Please make sure that you print out only the requested names for each year, not the data frame or all the names.  In general, printing out a large part of a dataset could potentially cause Rstudio to have compiling issues since there will be many pages of data to be printed out.


```{r} 
names2000.2003 |> arrange(name, by_group = T) |> filter(nrow(names2000.2003) == 20)

```


## More on baby names

To see how Jared used data to help pick out his son's name see [his talk](https://jaredlander.com/content/2019/11/BabyData.html) from the [2019 Washington DC R Conference](https://www.youtube.com/watch?v=CPNphrraLsw).

