---
title: "Precept 8 Exercises"
author: "Your Name"
date: ""
output:
  pdf_document:
    fig_caption: yes
    number_sections: no
    toc: no
    toc_depth: 3
  html_document:
    df_print: paged
    toc: no
    toc_depth: 3
geometry: margin=1.5in
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align="center", fig.height=4, fig.width=6, collapse=TRUE, comment=NA, prompt=FALSE, echo = TRUE, cache=TRUE, autodep=TRUE, tidy.opts=list(width.cutoff=63),tidy=TRUE)
options(width=63)
```

# Objectives

* Conducting hypothesis tests
* Understand regression outputs
* Learning functions: `dt(x, df, ...)`, `pt(q, df, ...)`, `qt(p, df, ...)`, `rt(n, df, ...)`
* t.test(x, y = NULL,
       alternative = c("two.sided", "less", "greater"),
       mu = 0, paired = FALSE, var.equal = FALSE,
       conf.level = 0.95, ...)

# Key Concepts

## t distribution

Let $t_{df}$ be the t-distribution with the degree of freedom $df$.  The following functions are related to X ~ $t_{df}$ random variables (the first letter of the function gives us a hint what the output is):

* dt(x, df, ...) -- returns **density** at X = x
* pt(q, df, ...) -- returns the **probabliity** P(X <= q)  
* qt(p, df, ...) -- Given p, returns the **quantile** q such that the P(X <= q) = p 
* rt(n, df, ...) -- Generate a **random sample** of size `n`, drawn from the $t_{df}$ distribution

## Use `t.test()` to construct a CI


`t.test(x = sampl, conf.level = 0.95)` creates a 95\% confidence interval to predict the population mean with the numbers in `sampl`.  Note that `sampl` has to be a vector.
       
       
# Exercise 1

(Hypothetical) One month before the presidential election in 2020 we take a simple random sample from all the active New Jersey (NJ) voters and ask them whom they will vote for.  There are 1000 people in the sample and 46% of them say that they will vote for the Democratic candidate, Joe Biden, and 43% say that they will vote for the Republican candidate, 
Donald Trump.  The percentages do not add up to 100% because there are other candidates too.  

We are interested in predicting which candidate is going to win the electoral college votes for New Jersey; i.e., we are interested in whether the percentage of the New Jersey active voters voting for Biden is greater than that for Trump.

## Part a

What is the parameter that we are trying to estimate here?

Answer:  Either of the following two will be an acceptable answer but please use the first version for the rest of the questions:

Proportion of New Jersey voters voting for Biden - Proportion of New Jersey voters voting for Trump

Proportion of New Jersey voters voting for Trump - Proportion of New Jersey voters voting for Biden 

>. The true proportion of the New Jersey active voters population voting for Biden over Trump (Proportion of New Jersey voters voting for Biden - Proportion of New Jersey voters voting for Trump)

## Part b

What range of the values for the parameter implies that Democrat wins the electoral college votes for New Jersey?

Answer: 

>. (Proportion of New Jersey voters voting for Biden - Proportion of New Jersey voters voting for Trump) > 0 

## Part c

What is the statistic that we can use to estimate the population parameter?

Answer:  

>. 46% - 43% = 3% (Proportion of New Jersey voters voting for Biden - Proportion of New Jersey voters voting for Trump in the sample) which is the statistic from the sample that we can use to estimate the population parameter highlighted in part a).

## Part d

Since we know the behavior of the sample mean random variable when the sample size is large and when the sample is a simple random sample, we would like to to formulate the problem in such a way that the parameter is the mean of a population and the statistic is the sample mean.  What number would you record for each person in the population?  

Hint: Suppose I have a simple random sample of 10 people.  Among them 5 said that they would vote for Biden, 4 for Trump and 1 for one of the other candidates.   What number will you record for each of these people in the sample so that the sample mean will be the statistic mentioned in part c?

Answer: 

>. Record a +1 for Biden voters, a -1 for Trump voters, and a 0 for voters of the other candidates. Then, I would calculate the mean for the sum of all of the voters based on who they voted for (the mean of these recorded numbers). 

## Part e

Construct a 90% C.I. to estimate the parameter in part a.  Do this part without using t.test().

Answer:  
```{r}
sampl <- rep(-1:1, times = c(430, 110, 460))
c(mean(sampl) - qt(p = .95, df = length(sampl) - 1)*sd(sampl)/sqrt(1000),
  mean(sampl) + qt(p = .95, df = length(sampl) - 1)*sd(sampl)/sqrt(1000))

```

The 90% CI is $$(-0.01911597,  0.07911597)$$

## Part f

Repeat Part f but now use t.test() (see Ch 5.3 on how to do this).  Check that you get the correct answer by comparing your answer for this part to that for the previous part.


Answer:  
```{r}
sampl <- rep(-1:1, times = c(430, 110, 460))
t.test(x = sampl, conf.level = .90)
# CI: -0.01911597  0.07911597

```



## Part g

Does your 90% CI cover zero?  Note that the sample margin is $\frac{460 - 430}{1000} = 3\%$.  Does having a CI that covers zero mean that you are more or less sure about that Biden will do better than Trump?

Answer: 

> Yes, the 90% CI does cover zero. Having a CI that covers zero means that we are less sure that Biden will do better than Trump even though a larger sample size (1000) means a smaller SE and a narrower CI (more precise). The CI is uncertain that the sample’s result generalizes to the population so we can't come to a conclusion on who will do better in the end.

## Part h

Let's call the parameter that we want to estimate $\mu$ and let's call the sample mean for our sample (of size 1000), $\bar{X}$.  What is the approximate distribution of $\frac{\bar{X} - \mu}{SE(\bar{X})}$ supposed to be?  Report the numeric value(s) for the parameter(s) for this distribution too.  Note that our sample of size 1000 is a simple random sample from the population of all the active NJ voters.

Answer:  

```{r}
pt(.9, df = length(sampl)-1)
# Numeric value for the parameter for the distribution: 0.8158314
```

>. The numeric value for the parameter of the approximate distribution using t.test is 0.8158314

## Part i

Test the following hypotheses:

$H_0$: The percentage of the New Jersey active voters supporting Biden is less than and equal to that supporting Trump.  
$H_1$: The percentage is greater for Biden

Report the standardized test statistic and the p-value for the test.

Answer:

```{r}
sampl <- rep(-1:1, times = c(430, 110, 460))
(.03-0)/(sd(sampl)/sqrt(1000))
# Standardized test statistic: 1.005608
1 - pt((.03-0)/(sd(sampl)/sqrt(1000)), df = length(sampl) - 1)
# p-value: 0.1574239

```

>. The standardized test statistic is 1.005608 and the p-value is about 0.1574239.

## Part j

According to your result in the previous part will you reject the null hypothesis at the 10% significant level?

Answer:  

>. No, I will not reject the null hypothesis because the p-value (~15.7%) is greater than the the significance level of 10% which means there is not enough evidence to reject the claim made.



### Question 2 (This question is the same as questions 3.d and 3.e in the last precept)

This dataset \emph{Newcomb.RData} includes data on the amount of time (in seconds) taken for a beam of light to travel from the Naval Observatory in Washington D.C. to the Washington Monument and back, a distance of 7.44373km.   The data was collected in 1880 by Simon Newcomb (working with Ablert Michelson).

```{r}

Newcomb =data.frame(
    Time = c(2.4828e-05, 2.4826e-05, 2.4833e-05, 2.4824e-05, 
             2.4834e-05, 2.4756e-05, 2.4827e-05, 2.4816e-05, 2.484e-05, 2.4798e-05,
             2.4829e-05, 2.4822e-05, 2.4824e-05, 2.4821e-05, 2.4825e-05, 2.483e-05,
             2.4823e-05, 2.4829e-05, 2.4831e-05, 2.4819e-05, 2.4824e-05, 2.482e-05,
             2.4836e-05, 2.4832e-05, 2.4836e-05, 2.4828e-05, 2.4825e-05, 2.4821e-05,
             2.4828e-05, 2.4829e-05, 2.4837e-05, 2.4825e-05, 2.4828e-05, 2.4826e-05,
             2.483e-05, 2.4832e-05, 2.4836e-05, 2.4826e-05, 2.483e-05, 2.4822e-05,
             2.4836e-05, 2.4823e-05, 2.4827e-05, 2.4827e-05, 2.4828e-05, 2.4827e-05,
             2.4831e-05, 2.4827e-05, 2.4826e-05, 2.4833e-05, 2.4826e-05, 2.4832e-05,
             2.4832e-05, 2.4824e-05, 2.4839e-05, 2.4828e-05, 2.4824e-05, 2.4825e-05,
             2.4832e-05, 2.4825e-05, 2.4829e-05, 2.4827e-05, 2.4828e-05, 2.4829e-05,
             2.4816e-05, 2.4823e-05), 
    Series = as.factor(
        c(1, 1, 1, 1, 1, 
          1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2,
          2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3, 3, 3,
          3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3))) 
str(Newcomb)
head(Newcomb)
```

#### Part a

We would like to construct a 90\% confidence intervals to predict the population standard deviation of *Time*.  Since we do not know what kind of distribution the sample SD follows and how it relates to the population SD, we will use Bootstrap sampling to help us get an idea of the distribution of the sample SD.  Use bootstrap sampling with 20,000 simulations to determine the 90\% confidence intervals for predicting the population standard deviation of *Time*.  Use `set.seed(672)`.

```{r}
set.seed(672)
# Create Bootstrap samples and take the mean for each of them
library(purrr)
boot.sd = map_dbl(1:20000, function(x){
              sd(sample(Newcomb$Time, size = length(Newcomb$Time), replace = T))
                                    }
                   )

# Calculate the deviations of Bootstrap samples
pop.deviation = boot.sd-sd(Newcomb$Time)
```




#### Part b

Make a density plot of the bootstrap sample SD's above. Does the distribution of the bootstrap sample SD's look Normal?

```{r}
ggplot(tibble(pop.deviation), mapping = aes(x = pop.deviation)) + 
  geom_histogram(mapping = aes(y = after_stat(density)), bins = 30, color = "white", fill = "steelblue") +
  geom_density(aes(color = "Sample data"),
               linewidth = 1.2) +
  stat_function(fun = dnorm, 
                args = list(mean = mean(pop.deviation), sd = sd(pop.deviation)), 
                mapping = aes(color = "Normal distribution"), 
                linewidth = 1.2) +
  scale_colour_manual(name = "Density Curve", values = c("Normal distribution" ="pink","Sample data" = "yellow"))

```

>. No, the distribution of the bootstrap SD's does not look Normal based on the alignment of the density curves.
